Analiza vremenskih nizova zasnovana na kompleksnim mrežama

### 2017-06-05 tex/diplomski.tex
E>

\documentclass[lmodern, utf8, diplomski]{fer}
\usepackage{booktabs}

\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\microtypecontext{spacing=nonfrench}
\usepackage{indentfirst}
\usepackage{booktabs}
\usepackage{tabularx}

@@

\chapter{Uvod}
  Classical statistical arbitrage methods take into account a pair of assets whose prices behave similarly during certain period of time.
  Similarity is measured by cointegration, correlation, or some other measure in order to find a moment in time when those assets' prices exceed what was statistically determined as highly confident range.
  When such opportunities present themselves, we can take advantage by predicting that prices will return once again to the confident range in the next time step, and do the trading in accordance with this prediction.

@@

@@

%  Let $\Omega$ be any set of entities.
  Neka je $\Omega$ skup općenitih dobara.
%  Preference relation $\succ$ defined over $\Omega \times \Omega$ is a strict weak ordering that describes the way humans prefer some entity over another.
  Relacija preferencije, označena sa $\succ$ i definirana nad $\Omega \times \Omega$, je strogi slabi uređaj koji odgovara načinu na koji ljudi preferiraju jedno dobro u odnosu na drugo.
  Između dva dobra $a, b \in \Omega$ relacija može, ali i ne mora postojati.
  Primjerice, $a$ može biti više preferirano u odnosu na $b$, ili $b$ u odnosu na $a$, ali moguća je situacija gdje su oba dobra podjednako preferirana.
  U tom slučaju radi se o indiferentnosti između $a$ i $b$, i to se označava kao $a \sim b$.
%  This relation is specific in that it is ($\forall x, y, z \in \Omega$):
  Ova relacija specifična je po tome što je ($\forall x, y, z \in \Omega$):
  \begin{itemize}
%    \item \textit{irreflexive}: every entity $x$ is not preferrable over itself, %($\neg \left( x \succ x \right)$),
    \item \textit{irrefleksivna}: $\neg \left( x \succ x \right)$ --- za nijedno dobro ne vrijedi da je više preferirano od samog sebe,
%    \item \textit{asymmetrical}: if $x$ is preferrable over some $y$, then $y$ is not preferrable over $x$, %($x \succ y \Rightarrow \neg \left( y \succ x \right)$),
    \item \textit{asimetrična}: $x \succ y \Rightarrow \neg \left( y \succ x \right)$ --- ako je $x$ više preferirano od $y$, onda $y$ nije više preferirano od $x$,
%    \item \textit{transitive}: if $x$ is preferrable over $y$, and $y$ is preferrable over $z$, then $x$ is also preferrable over $z$, %($x \succ y \wedge y \succ z \Rightarrow x \succ z$),
    \item \textit{tranzitivna}: $x \succ y \wedge y \succ z \Rightarrow x \succ z$ --- ako je $x$ više preferirano od $y$, te $y$ više preferirano od $z$, tada je i $x$ više preferirano od $z$,
%    \item \textit{transitive in incomparability} (noting that $x$ and $y$ may be \textit{incomparable}, i.e. neither $x$ is preferrable over $y$, nor $y$ is preferrable over $x$): if $x$ is incomparable with $y$, and $y$ is incomparable with $z$, then $x$ is also incomparable with $z$.
    \item \textit{tranzitivna u indiferentnosti}: $x \sim y \wedge y \sim z \Rightarrow x \sim z$ --- ako je $x$ podjednako preferirano kao i $y$, te $y$ podjednako preferirano kao i $z$, tada je i $x$ podjednako preferirano kao i $z$.
  \end{itemize}

%  We naturally assume this kind of relation we describe relationships among the assets.
%  Determining that some asset is preferred over another is easier than assigning a preference rank to each asset individually, especially for a larger number of assets.
%  However, the latter is more useful for decision making, and therefore it is desirable to find a way of sorting assets in the order of preference.
  Ovakva vrsta relacije prirodno opisuje odnose među raznim imovinama, npr. dionica $A$ u nekom trenutku može biti više preferirana od dionice $B$.
  Razlog tome je što je za čovjeka lakše ocijeniti odnos (više, manje, jednako preferirano) između svakih dviju dionica, nego što je pridijeliti individualnu mjeru preferencije za svaku dionicu, pogotovo ako se radi o velikom broju dionica.
  No ipak, u svrhu kreiranja portfelja više se isplati imati individualnu mjeru preferencije za svaku dionicu.
  Stoga je poželjno pronaći način da se iz relacije preferencije dobiju individualne mjere preferencije.

%  Utility function $U\colon \Omega \to \mathbb{R}$ is mapping from entities to real numbers, in such way that order of the mapping corresponds to the preference order of the entities, i.e. $\forall x, y \in \Omega, U(x) > U(y) \Rightarrow x \succ y$.
%  One such mapping is obtained by using potential method that is described later in this paper.
%  In addition to ordering of the entities, utility function also provides the intensity of preference for a particular entity, hence it is more informative when it comes to the decision making.
  Funkcija korisnosti $U\colon \Omega \to \mathbb{R}$ je preslikavanje iz skupa dobara u skup realnih brojeva, na način da poredak preslikanih realnih brojeva odgovara poretku dobara prema individualnoj preferenciji.
  Jedno takvo preslikavanje dobije se korištenjem metode potencijala koja je opisana u poglavlju \ref{sec:metpot}.
  Uz to što određuje poredak dobara premda individualnoj preferenciji, funkcija korisnosti također unosi i mjeru intenziteta preferencije za neko dobro, što znači da nosi više informacija od same relacije.

  \section{Graph of preference flow}

  Graph of preference flow is a weighted directed graph without multiple edges and loops, whose nodes represent entities, edges represent preference for one entity over another, and edge weights correspond to the strength of the preferences.
  If an edge between nodes is missing, it is considered that neither entity is preferable over another (incomparability of entities).
  The graph as a whole describes preference flow among the entities.
  An example of graph is shown on Fig. \ref{fig:graph}.

  \begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth]{graphics/graph.pdf}
    \caption{(TODO: put image with weights on edges) An example of a graph of preference flow. Asset number and calculated preference are inscribed in each node, and preference flows are shown on edges.}
    \label{fig:graph}
  \end{figure}

  Construction of the graph is based on a statistical arbitrage method.
  An edge going from node $i$ to node $j$ with weight $w_{i,j}$ will be present in the graph if and only if assets represented by nodes $i$ and $j$ have demonstrated similar behavior during the past period, but have suddenly diverged at the moment, as determined per statistical measures.
  Weight $w_{i,j}$ corresponds to the magnitude of this divergence.  
  A detailed description of used statistical measures the procedure is given later in \ref{sub:creating-graph}.

  Connections in this graph impose a preference relation among entities that are presented by the graph, in the way that an edge that goes from node $A$ to node $B$ means that $A$ is preferred over $B$.
  It is the case that neither node is in relation with itself (irreflexivity), and that no multiple connections are allowed between two nodes (implies asymmetry).
  However, problems arise with the aforementioned properties of transitivity, and transitivity in incomparability, which may not hold for an arbitrarily constructed instance of the graph.
  This imposed preference relation should preferably be in compliance with all the aforementioned properties, but when it comes to larger number of entities, it may become infeasible to construct a graph of such qualities directly.
  Instead of aiming at a consistent preference relation, we devise a consistency measure that describes similarity between the original graph and its nearest consistent reconstruction, and use it as an additional parameter in decision making.

@@

  \section{Potential method}
  \label{sec:metpot}
  From previously obtained graph it is possible to tell which pair of assets has the highest preference flow.
  However, it is not yet possible to directly tell which are the most or least preferrable assets, or obtain the measure of preference for individual assets.
  To calculate preferences for each node in the graph, we use the potential method\cite{caklovic}.
  The potential of a node corresponds to difference in amount of flow going in and out of the node.

@@

  \section{Creating the graph of preference flow}
  \label{sub:creating-graph}
  Using the obtained $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ it is now possible to create a graph of preference flow among assets for each time step $t$.
  Considering one time step $t$, we find all such pairs of assets $(i,j)$ that satisfy:
  \begin{equation}
  \label{eq:thresh}
  \left| c_{i,j}^{(t)} - m_{i,j}^{(t)} \right| > \alpha \cdot d_{i,j}^{(t)},
  \end{equation}
  i.e. current log price difference is at least $\alpha$ deviations distant from mean value of the past time window.
  An illustration is shown on Fig. \ref{fig:devmag}.
  %  Parameter $\alpha$ determines how many pairs of assets should constitute the graph at current time step.
  Afterwards, for each observed pair $(i,j)$ that exceeds the threshold we add into graph vertices $i$ and $j$, with a weighed edge of weight $w_{i,j}^{(t)}$ going from $i$ to $j$.
  Weight $w_{i,j}^{(t)}$ is obtained as:
  \begin{equation}
  \label{eq:weight}
  w_{i,j}^{(t)} = \left. \left(c_{i,j}^{(t)} - m_{i,j}^{(t)}\right) \middle/ d_{i,j}^{(t)} \right..
  \end{equation}

@@

  \section{Choosing assets from graph}
  We obtain preference for each asset via the potential method, as described earlier in \ref{sub:potential}.
  By obtaining the measure of preference for each asset it is possible to pick assets for the portfolio.
  The most preferred assets should be bought while the least preferred should be short-sold if possible.

@@

  \chapter{Results}

>>>>

\documentclass[lmodern, utf8, diplomski, numeric]{fer}
\usepackage{booktabs}

\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\microtypecontext{spacing=nonfrench}
\usepackage{indentfirst}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{rotating}

@@

\chapter{Uvod}
%  Classical statistical arbitrage methods take into account a pair of assets whose prices behave similarly during certain period of time.
%  Similarity is measured by cointegration, correlation, or some other measure in order to find a moment in time when those assets' prices exceed what was statistically determined as highly confident range.
%  When such opportunities present themselves, we can take advantage by predicting that prices will return once again to the confident range in the next time step, and do the trading in accordance with this prediction.
  Klasične metode statističke arbitraže uzimaju u obzir parove vrijednosnica čije cijene se ponašaju slično tijekom određenom vremenskog perioda.
  Sličnost se mjeri kointegracijom, korelacijom, ili nekom drugom mjerom, s ciljem pronalaska trenutka kada te cijene izlaze van statistički utvrđenog intervala visoke pouzdanosti.
  Takve prilike mogu se iskoristiti predviđanjem da će se cijene u idućem trenutku ponovno vratiti unutar intervala visoke pouzdanosti, te se u skladu s tim predviđanjem može provesti trgovanje.

@@

  \section{Statistička arbitraža zasnovana na standardnoj devijaciji}

@@

%  Let $\Omega$ be any set of entities.
  Neka je $\Omega$ skup općenitih dobara.
%  Preference relation $\succ$ defined over $\Omega \times \Omega$ is a strict weak ordering that describes the way humans prefer some entity over another.
  Relacija preferencije, označena sa $\succ$ i definirana nad $\Omega \times \Omega$, je strogi slabi uređaj koji odgovara načinu na koji ljudi preferiraju jednu vrijednosnicu u odnosu na drugu.
  Između dvaju dobara $a, b \in \Omega$ relacija može, ali i ne mora postojati.
  Primjerice, $a$ može biti više preferirano u odnosu na $b$, ili $b$ u odnosu na $a$, ali moguća je situacija gdje su oba dobra podjednako preferirane.
  U tom slučaju radi se o indiferentnosti između $a$ i $b$, i to se označava kao $a \sim b$.
%  This relation is specific in that it is ($\forall x, y, z \in \Omega$):

  Ova relacija specifična je po tome što je ($\forall x, y, z \in \Omega$):
  \begin{itemize}
%    \item \textit{irreflexive}: every entity $x$ is not preferrable over itself, %($\neg \left( x \succ x \right)$),
    \item \textit{irrefleksivna}: $\neg \left( x \succ x \right)$ --- za nijedno dobro ne vrijedi da je više preferirano od samog sebe,
%    \item \textit{asymmetrical}: if $x$ is preferrable over some $y$, then $y$ is not preferrable over $x$, %($x \succ y \Rightarrow \neg \left( y \succ x \right)$),
    \item \textit{asimetrična}: $x \succ y \Rightarrow \neg \left( y \succ x \right)$ --- ako je $x$ više preferirano od $y$, onda $y$ nije više preferirano od $x$,
%    \item \textit{transitive}: if $x$ is preferrable over $y$, and $y$ is preferrable over $z$, then $x$ is also preferrable over $z$, %($x \succ y \wedge y \succ z \Rightarrow x \succ z$),
    \item \textit{tranzitivna}: $x \succ y \wedge y \succ z \Rightarrow x \succ z$ --- ako je $x$ više preferirano od $y$, te $y$ više preferirano od $z$, tada je i $x$ više preferirano od $z$,
%    \item \textit{transitive in incomparability} (noting that $x$ and $y$ may be \textit{incomparable}, i.e. neither $x$ is preferrable over $y$, nor $y$ is preferrable over $x$): if $x$ is incomparable with $y$, and $y$ is incomparable with $z$, then $x$ is also incomparable with $z$.
    \item \textit{tranzitivna po indiferentnosti}: $x \sim y \wedge y \sim z \Rightarrow x \sim z$ --- ako je $x$ podjednako preferirano kao i $y$, te $y$ podjednako preferirano kao i $z$, tada je i $x$ podjednako preferirano kao i $z$.
  \end{itemize}

%  We naturally assume this kind of relation we describe relationships among the assets.
%  Determining that some asset is preferred over another is easier than assigning a preference rank to each asset individually, especially for a larger number of assets.
%  However, the latter is more useful for decision making, and therefore it is desirable to find a way of sorting assets in the order of preference.
  Ovakva vrsta relacije prirodno opisuje odnose među različitim vrijenosnicama, npr. dionica $A$ u nekom trenutku može biti više preferirana od dionice $B$.
  Razlog tome je što je za čovjeka lakše ocijeniti odnos (više, manje, jednako preferirano) između svakog para vrijednosnica, nego pridijeliti svakoj vrijednosnici individualnu mjeru preferencije, pogotovo ako se radi o velikom broju vrijednosnica.
  No ipak, u svrhu konstruiranja portfelja korisnije je posjedovati individualnu mjeru preferencije za svaku vrijednosnicu.
  Stoga je poželjno pronaći način da se iz relacije preferencije dobiju individualne mjere preferencije.

%  Utility function $U\colon \Omega \to \mathbb{R}$ is mapping from entities to real numbers, in such way that order of the mapping corresponds to the preference order of the entities, i.e. $\forall x, y \in \Omega, U(x) > U(y) \Rightarrow x \succ y$.
%  One such mapping is obtained by using potential method that is described later in this paper.
%  In addition to ordering of the entities, utility function also provides the intensity of preference for a particular entity, hence it is more informative when it comes to the decision making.
  Funkcija korisnosti $U\colon \Omega \to \mathbb{R}$ je preslikavanje iz skupa dobara u skup realnih brojeva, na način da poredak preslikanih realnih brojeva odgovara poretku dobara prema individualnoj preferenciji, tj. vrijedi $\forall x, y \in \Omega, U(x) > U(y) \Leftrightarrow x \succ y$.
  Jedno takvo preslikavanje ostvaruje se korištenjem metode potencijala koja je opisana u poglavlju \ref{sec:metpot}.
  Uz to što određuje poredak dobara prema individualnoj preferenciji, funkcija korisnosti također unosi i mjeru intenziteta preferencije prema nekoj vrijednosnici, što znači da nosi više informacija od same relacije.

  \section{Graf toka preferencija}

%  Graph of preference flow is a weighted directed graph without multiple edges and loops, whose nodes represent entities, edges represent preference for one entity over another, and edge weights correspond to the strength of the preferences.
%  If an edge between nodes is missing, it is considered that neither entity is preferable over another (incomparability of entities).
%  The graph as a whole describes preference flow among the entities.
%  An example of graph is shown on Fig. \ref{fig:graph}.
  Graf toka preferencija je težinski usmjereni graf, bez višestrukih bridova i petlji.
  Njegovi čvorovi predstavljaju vrijednosnice, usmjereni bridovi preferenciju jedne vrijednosnice nad drugom, a težine bridova odgovaraju jačini preferencije.
  Ukoliko između dva čvora nedostaje brid, smatra se da su pripadne vrijednosnice podjednako preferirane (indiferentnost u odlučivanju).
  Graf kao cjelina opisuje tok preferencija među vrijednosnicama.
  Primjer grafa prikazan je na slici \ref{fig:graph}.

  \begin{figure}[h]
    \centering
    \includegraphics[width=0.5\columnwidth]{graphics/graph.pdf}
    \caption{Primjer grafa toka preferencija. Na bridovima su prikazane jačine preferencija jedne vrijednosnice u odnosu na drugu, a u čvorove su upisane izračunate individualne mjere preferencija za pripadnu vrijednosnicu.}
    \label{fig:graph}
  \end{figure}

%  Construction of the graph is based on a statistical arbitrage method.
%  An edge going from node $i$ to node $j$ with weight $w_{i,j}$ will be present in the graph if and only if assets represented by nodes $i$ and $j$ have demonstrated similar behavior during the past period, but have suddenly diverged at the moment, as determined per statistical measures.
%  Weight $w_{i,j}$ corresponds to the magnitude of this divergence.  
%  A detailed description of used statistical measures the procedure is given later in \ref{sub:creating-graph}.
  Konstrukcija grafa temelji se na metodi statističke arbitraže.
  Brid koji ide od čvora $i$ do čvora $j$ s težinom $w_{i,j}$ je prisutan u grafu ako i samo ako su se dvije pripadne vrijednosnice $i$ i $j$ ponašale slično tijekom proteklog vremenskog perioda, ali su prema određenim statističkim mjerama trenutačno razdvojile.
  Težina $w_{i,j}$ opisuje magnitudu ovog razdvajanja.
  Detaljan opis korištenih statističkih mjera dan je u odjeljku \ref{sub:creating-graph}.

%  Connections in this graph impose a preference relation among entities that are presented by the graph, in the way that an edge that goes from node $A$ to node $B$ means that $A$ is preferred over $B$.
%  It is the case that neither node is in relation with itself (irreflexivity), and that no multiple connections are allowed between two nodes (implies asymmetry).
%  However, problems arise with the aforementioned properties of transitivity, and transitivity in incomparability, which may not hold for an arbitrarily constructed instance of the graph.
%  This imposed preference relation should preferably be in compliance with all the aforementioned properties, but when it comes to larger number of entities, it may become infeasible to construct a graph of such qualities directly.
%  Instead of aiming at a consistent preference relation, we devise a consistency measure that describes similarity between the original graph and its nearest consistent reconstruction, and use it as an additional parameter in decision making.
  Veze u ovom grafu na jedan način nameću relaciju preferencije među vrijednosnicama koje su opisane grafom, na način da brid koji ide iz čvora $i$ u čvor $j$ pokazuje da je vrijednosnica $i$ više preferirana od vrijednosnice $j$.
  Kako ne graf sadrži petlje, tj. nijedna vrijednosnica nije u relaciji sama sa sobom, ispunjeno je svojstvo irrefleksivnosti; a kako nema višestrukih bridova među čvorovima vrijedi i svojstvo asimetričnosti.
  Ipak, problemi se pojavljuju kod prethodno spomenutih svojstava tranzitivnosti, i tranzitivnosti po indiferentnosti, koja ne moraju uvijek vrijediti za proizvoljno konstruiran graf.
  Poželjno bi bilo da graf nameće relaciju preferencije koja zadovoljava sva četiri prethodno navedena svojstva, no kada se radi o većem broju vrijednosnica, konstruiranje grafa koji posjeduje takva svojstva postaje nepraktično.
  Umjesto postizanja konzistentnosti sa svojstvima relacije preferencije, upotrijebljena je mjera konzistentnosti koja opisuje koliko je neki graf sličan sa svojom najbližom konzistentnom rekonstrukcijom, te se ona koristi kao dodatan parametar pri konstruiranju portfelja.
  Opis mjere konzistentnosti te način dobivanja najbliže konzistentne rekonstrukcije grafa dan je u odjeljku \ref{sec:metpot}.

@@

  \section{Metoda potencijala}
  \label{sec:metpot}
  From previously obtained graph it is possible to tell which pair of assets has the highest preference flow.
  However, it is not yet possible to directly tell which are the most or least preferrable assets, or obtain the measure of preference for individual assets.
  To calculate preferences for each node in the graph, we use the potential method\cite{caklovic}.
  The potential of a node corresponds to difference in amount of flow going in and out of the node.

@@

  \section{Konstrukcija grafa toka preferencija}
  \label{sub:creating-graph}
  Using the obtained $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ it is now possible to create a graph of preference flow among assets for each time step $t$.
  Considering one time step $t$, we find all such pairs of assets $(i,j)$ that satisfy:
  \begin{equation}
  \label{eq:thresh}
  \left| c_{i,j}^{(t)} - m_{i,j}^{(t)} \right| > \alpha \cdot d_{i,j}^{(t)},
  \end{equation}
  i.e. current log price difference is at least $\alpha$ deviations distant from mean value of the past time window.
  An illustration is shown on Fig. \ref{fig:devmag}.
  %  Parameter $\alpha$ determines how many pairs of assets should constitute the graph at current time step.
  Afterwards, for each observed pair $(i,j)$ that exceeds the threshold we add into graph vertices $i$ and $j$, with a weighed edge of weight $w_{i,j}^{(t)}$ going from $i$ to $j$.
  Weight $w_{i,j}^{(t)}$ is obtained as:
  \begin{equation}
  \label{eq:weight}
  w_{i,j}^{(t)} = \left. \left(c_{i,j}^{(t)} - m_{i,j}^{(t)}\right) \middle/ d_{i,j}^{(t)} \right..
  \end{equation}

@@

  \section{Konstrukcija portfelja prema grafu toka preferencija}
  We obtain preference for each asset via the potential method, as described earlier in \ref{sub:potential}.
  By obtaining the measure of preference for each asset it is possible to pick assets for the portfolio.
  The most preferred assets should be bought while the least preferred should be short-sold if possible.

@@

  \chapter{Rezultati}
